---
title: "R notebook: Model Specification"
author: "Dr Sergey Kushnarev"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
---

## Steps in Time Series Analysis            

|   |   |   |   |
|:---:|:---:|:---:|:---:|
| **Step 1: Model Specification** |  Step 2: Parameter Estimation  | Step 3: Model Checking | Step 4: Forecasting |

# Model Specification

This R notebook is based on Chapter 6, Cryer and Chan (2008), *Time Series Analysis with Applications in R*. Springer.

We have developed a large class of parametric models for both stationary and nonstationary
time series -- the ARIMA models. We now begin our study and implementation
of statistical inference for such models. The subjects of the next three chapters, respectively,
are:

1. How to choose  $p, d$, and $q$ for a given series? (Chapter 6);
2. How to estimate $\phi_i,\theta_j$ for a specific ARIMA(p,d,q)? (Chapter 7);
3. Does the model fit well? (Chapter 8).
4. Forecast the future. (Chapter 9).

Our overall strategy will first be to decide on reasonable preliminary values
for $p$, $d$, and $q$ (this notebook). 

We will fit a model to the data (Chapter 7) and then check the residuals for
autocorrelation and normality (Chapter 8). If the residuals are not white noise, we will try to 
improve the model by changing the values of $p$, $d$, and $q$.

```{r, message=FALSE}
library(TSA)
```

## Specification of simulated MA(q) processes

```{r}
# Load the data
data(ma1.1.s)
# ACF of the series
acf(ma1.1.s,
    xaxp=c(0,20,10),
    main = "ACF of MA(1) with theta=0.9")
```

```{r}
acf(ma1.1.s,
    ci.type='ma',
    xaxp=c(0,20,10),
    main = "Alternative bounds for the ACF of MA(1) with theta=0.9")
```

```{r}
data(ma1.2.s)
acf(ma1.2.s,
    xaxp=c(0,20,10),
    main = "ACF of MA(1) with theta=-0.9")

data(ma2.s)
acf(ma2.s,
    xaxp=c(0,20,10),
    main = "ACF of MA(2) with theta1=1, theta2=-0.6")
acf(ma2.s,
    ci.type='ma',
    xaxp=c(0,20,10),
    main = "Alternative bounds for the ACF of MA(2) with theta1=1, theta2=-0.6")
```

## Specification of the simulated AR(p) processes

```{r}
data(ar1.s) 
acf(ar1.s,
    xaxp=c(0,20,10),
    main = "ACF of AR(1) with phi=0.9") 
pacf(ar1.s,xaxp=c(0,20,10),
      main = "PACF of AR(1) with phi=0.9")
```

```{r}
data(ar2.s)
acf(ar2.s,
    xaxp=c(0,20,10),
    main = "ACF of AR(2) with phi1=1.5, phi2=-0.75")
pacf(ar2.s,
     xaxp=c(0,20,10),
     main = "PACF of AR(2) with phi1=1.5, phi2=-0.75")
```

## Specification of the simulated ARMA(p,q) processes

```{r}
data(arma11.s)
plot(arma11.s, type='o',ylab=expression(Y[t]),
    main = "ARMA(1,1) with phi=0.6, theta=-0.3")
acf(arma11.s,xaxp=c(0,20,10),
    main = "ACF of ARMA(1,1) with phi=0.6, theta=-0.3")
pacf(arma11.s,xaxp=c(0,20,10),
     main = "PACF of ARMA(1,1) with phi=0.6, theta=-0.3")
```


```{r}
# In an R Notebook, simply run this code chunk:

# 1) Create an empty matrix filled with "X":
n_ar <- 8   # AR orders from 0 to 7
n_ma <- 14  # MA orders from 0 to 13
eacf_mat <- matrix("X", nrow = n_ar, ncol = n_ma)

# 2) Fill with "0" below the main diagonal:
#    For AR>0, columns at or past that AR index become "0".
for (i in 2:n_ar) {                  # i=2 means AR=1, i=3 means AR=2, etc.
  for (j in i:n_ma) {                # j runs from "i" through 14
    eacf_mat[i, j] <- "0"
  }
}

# 3) Put the special "0*" mark (the star at AR=1, MA=0):
eacf_mat[2, 2] <- "0*"

# 4) Label rows and columns:
rownames(eacf_mat) <- paste0("AR=", 0:(n_ar-1))
colnames(eacf_mat) <- paste0("MA=", 0:(n_ma-1))

# 5) Display the table nicely (using knitr::kable or similar):
library(knitr)
kable(eacf_mat, align = "c", caption = "Theoretical EACF for ARMA(1,1)")
```


```{r}
# Compute EACF
eacf(arma11.s)
```

## Nonstationarity





## Overdifferencing
```{r}
# Differencing the random walk twice
z <- rnorm(60)
rw <- ts(cumsum(z))
plot(diff(diff(rw)))
# ACF and PACF of the twice-differenced series
acf(diff(diff(rw)), 
    xaxp=c(0,22,11),
    ci.type='ma',
    main="Sample ACF of the overdifferenced random walk")
acf(diff(rw), 
    xaxp=c(0,22,11),
    main="Sample ACF of the correctly differenced random walk")
```

## Detecting nonstationarity: Augmented Dickey-Fuller test

```{r, warning=FALSE}
# Augmented Dickey-Fuller test
# Unit root test of the random walk and its first difference
library(tseries)
adf.test(rw)
adf.test(diff(rw))
```


## Model Selection: ARMA Subsets based on BIC

Schwarz *Bayesian Information Criterion* (BIC)
$$
BIC = – 2\log(maximum likelihood) + k\log(n)
$$
where $k$ is the number of parameters in the model and $n$ is the number of observations.

Minimizing the BIC is equivalent to maximizing the likelihood of the model while penalizing for the number of parameters.

```{r}
# ARMA Subsets
set.seed(92397)
yt <- arima.sim(model=list(ar=c(rep(0,11),0.8),
                            ma=c(rep(0,11),0.7)),n=120)
res <- armasubsets(y=yt,nar=14,nma=14,
                    y.name='yt',ar.method='ols')
plot(res)
```



```{r}
# Load the oil price data
data(oil.price)
# Plot the oil price series
acf(as.vector(diff(log(oil.price))),xaxp=c(0,22,11))
pacf(as.vector(diff(log(oil.price))),xaxp=c(0,22,11))
```

## Random Walk

```{r}
# Generate a random walk
set.seed(439)
z <- rnorm(60)
rw <- ts(cumsum(z))
# Plot the random walk
plot(rw, main="Random Walk")
# Unit root test of the random walk and its first difference
adf.test(rw)
adf.test(diff(rw))
# ACF, PACF of the Random Walk
acf(rw, main="ACF of Random Walk")
pacf(rw, main="PACF of Random Walk")
```

After differencing
$$
W_t=\nabla Y_t = Y_t - Y_{t-1}.
$$


```{r}
# ACF, PACF of the differenced Random Walk
acf(diff(rw), main="ACF of differenced Random Walk")
pacf(diff(rw), main="PACF of differenced Random Walk")
```

## The Los Angeles Annual Rainfall Series

```{r}
# Load the data
data(larain)
plot(larain)
# Stabilize the variance
qqnorm(log(larain),
       main = "Normal Q-Q Plot of log(larain)") 
qqline(log(larain))
```

Displaying sample autocorrelations

```{r}
acf(log(larain),
    xaxp=c(0,20,10),
    main = "ACF of log(larain)")
mean(log(larain))
sd(log(larain))
```

## The Chemical Process Color Property Series

```{r}
# Load the data
data(color)
plot(color)
# Plot ACF and PACF
acf(color,ci.type='ma', lag.max=20, main="ACF of color")
pacf(color, lag.max=20, main="PACF of color")
```

## The Annual Abundance of Canadian Hare Series

```{r}
# Load the data
data(hare)
plot(hare)
# Box-Cox transformation
BoxCox.ar(hare)
# Plot ACF and PACF of the transformed series
acf(hare^.5, main="ACF of hare^.5")
pacf(hare^.5, main="PACF of hare^.5")
```

## The Oil Price Series

```{r}
library(tseries)
# Load the data
data("oil.price")
plot(oil.price)
# Plot ACF and PACF
acf(as.vector(oil.price),
    xaxp=c(0,24,12),
    main="ACF of oil.price")
pacf(as.vector(oil.price), 
    xaxp=c(0,24,12),
    main="PACF of oil.price")
```

```{r}
# Log transformation
plot(log(oil.price))
# Differencing the log-transformed series
plot(diff(log(oil.price)))
acf(diff(as.vector(log(oil.price))),xaxp=c(0,24,12))
pacf(diff(as.vector(log(oil.price))),xaxp=c(0,24,12))
```

```{r}
# ADF test: to test for stationarity of log-transformed series
adf.test(log(oil.price))
adf.test(diff(log(oil.price)))
```

```{r}
res <- armasubsets(y=diff(log(oil.price)),nar=7,nma=7,y.name='test', ar.method='ols')
plot(res, main = "ARMA Subsets")
```
